---
title: "Malaria cases vs rainfall in Loreto Peru"
output: html_document
---

## Malaria cases vs rainfall in Loreto Peru

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(lubridate)

theme_set(theme_minimal())
```

## Import and wrangle data
The raw malaria case data come as one record for each case, tagged by vector, administrative unit and epidemiological week. We will aggregate the data to weekly counts of cases in each administrative unit. Additionally, we will limit our analysis to one vector: P. falciparum.

```{r malaria data import and wrangle}
#Import all the data
malaria_data <- read.csv(
  file=here("data", "raw", "datos_abiertos_vigilancia_malaria.csv"),
  stringsAsFactors = TRUE,
  colClasses = c('ubigeo'='factor')
  ) %>%
  #Filter out records with missing location
  filter(ubigeo != '999999')  %>% 
  #Create a column of the first 4 characters of ubigeo
  mutate(ubigeo_4 = substr(ubigeo, 1, 4))

## Create a lookup table for the administrative units
admin_units <- malaria_data %>%
  group_by(provincia, ubigeo_4) %>%
  summarise() 

## Wrangle the malaria data
malaria_data <- malaria_data %>%
  
  #Create a date from the year and week number
  mutate(
    date = ymd(paste0(ano, "-01-01")) + weeks(semana - 1),
    month = month(date)
    ) %>%
  
  #Filter for P. falciparum
  filter(enfermedad == "MALARIA P. FALCIPARUM") %>%
  
  #Group by date and administrative unit and count cases
  group_by(month, ano, ubigeo_4) %>%
  summarise(cases = n()) %>% 
  
  #Pivot to wide format on provincia
  pivot_wider(
    names_from = ubigeo_4, 
    values_from = cases, 
    values_fill = 0) %>% 
  
  #Re-add date column
  mutate(date = ymd(paste(ano, month, "01")))

## Add missing dates (months)
#Create a data frame with all the dates
all_dates <- data.frame(
  date = seq.Date(
    min(malaria_data$date), 
    max(malaria_data$date), 
    by = "month")
  )

#Join the data frame with the data
malaria_data <- left_join(all_dates, malaria_data,by='date') %>% 
  #Update missing date components
  mutate(
    ano = year(date),
    month = month(date),
  ) %>% 
  #Fill missing cases with 0
  replace(is.na(.), 0) %>%
  #Pivot longer
  pivot_longer(
    cols = -c(date, month, ano),
    names_to = "ubigeo_4",
    values_to = "cases"
    ) %>%
  #Join with admin_units
  left_join(admin_units, by = "ubigeo_4") %>% 
  #Rename ano column to year
  rename(year = ano)

#Clean up
remove (all_dates)

```

### Visualizations of malaria data
```{r malaria visualizations}
#Set the province
the_province = 'LORETO'

#Filter records for the province
plot_data <- malaria_data %>%
  filter(provincia == the_province)

#Line plot of monthly cases
plot_data %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line() +
  labs(
    title = paste("Malaria cases by month:",the_province),
    x = "Date",
    y = "Cases"
    ) 

#Boxplots by month
plot_data %>%
  ggplot(aes(x = factor(month,labels=month.abb), y = cases)) +
  geom_boxplot() +
  labs(
    title = paste("Malaria cases by month:",the_province),
    x = "Month",
    y = "Cases"
    ) 

#Find peaks in case rates
plot_data %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line() +
  geom_smooth() +
  labs(
    title = paste("Malaria cases by month:",the_province),
    x = "Date",
    y = "Cases"
    ) 

```

## Rainfall data
The rainfall data come as one record for each day, tagged by station. We will aggregate the data to weekly sums of rainfall in each station.

```{r rainfall data import and wrangle}
#Read in the rainfall data
rainfall_data <- read.csv(
  file=here("data", "raw", "PERU_LDAS.csv"),
  stringsAsFactors = TRUE,
  colClasses = c('ubigeo'='factor')
  ) %>% 
  
  #Subset columns for speed
  select(ubigeo:Rainfall.mm.) %>%
  
  #Update/create columns
  mutate(
    #Convert EpiweekStartDate to date
    EpiweekStartDate = ymd(EpiweekStartDate),
    #Create a month column
    month = month(EpiweekStartDate),
    #Create a ubigeo_4 column
    ubigeo_4 = substr(ubigeo, 1, 4)
    ) %>%
  #Compute monthly averages
  group_by(ubigeo_4, year, month) %>% 
  summarise(rainfall = mean(Rainfall.mm.)) 

#Join the rainfall to the malaria data
joined_data <- left_join(
  malaria_data, 
  rainfall_data, 
  by = c("ubigeo_4", "year", "month")
  ) %>% 
  #Filter for records after 2000
  filter(year > 2000) 
  
```

